---
title: "NYPD_Shooting"
author: "Henry Clay Ver Valen"
date: "5/21/2021"
output:
  pdf_document: default
  html_document: default
---

```{r load-packages, include=FALSE}
library(tidyverse)
library(lubridate)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, read in the NYPD Shooting incident data. The CSV file is located in ./data (use setwd to set your
working directory to the same folder as this Rmd file is located) and the original file can 
be downloaded at <https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic>.

You will need tidyverse and lubridate
install.packages("tidyverse")
install.packages("lubridate")
library(tidyverse)
library(lubridate)


```{r import_NYPD_data, message = FALSE}
shooting_data <- read_csv("./data/NYPD_Shooting_Incident_Data__Historic_.csv")
```

Now eliminate INCIDENT_KEY and all columns after VIC_RACE and convert OCCUR_DATE to a date data type
```{r tidy_shooting_data}
shooting_data <- shooting_data %>% 
  select(OCCUR_DATE:VIC_RACE) %>% 
  mutate(OCCUR_DATE = mdy(OCCUR_DATE))
```

Show summary of the data
```{r show_summary}
summary(shooting_data)
```

The visualizations I will be using do not require any filtering of missing values,but if it did I could do it with:
```{r remove_missing_data}
shooting_data_no_missing <- shooting_data %>% 
  filter(PERP_AGE_GROUP != "NA" & PERP_AGE_GROUP != "UNKNOWN" & PERP_SEX != "NA" & 
           PERP_RACE != "NA" & VIC_AGE_GROUP != "NA" & VIC_AGE_GROUP != "UNKNOWN" & 
           VIC_SEX != "NA" & VIC_RACE != "NA")
```

Group the data by month for both murders and shootings for the first visualization
```{r group_by_month}
month_group <- shooting_data %>% group_by(month=format(floor_date(OCCUR_DATE, "month"), format="%m")) %>%
  summarize(murders = sum(STATISTICAL_MURDER_FLAG, na.rm = TRUE), shootings = length(STATISTICAL_MURDER_FLAG)) %>%
  ungroup()
```
  
First visualization - Shootings and Murders by Month
``` {r first_viz}
month_group %>% 
  ggplot(aes(x = month, y = murders)) + 
  geom_point(aes(color = "murders")) + 
  geom_point(aes(y = shootings, color = "shootings")) + 
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "Shootings and Murders by Month", y = NULL)
```

Group the data by borough and year for murder and shootings for the second visualization
```{r group_by_boro_and_year}
boro_group <- shooting_data %>% group_by(BORO, year=year(OCCUR_DATE)) %>%
  summarize(murders = sum(STATISTICAL_MURDER_FLAG, na.rm = TRUE), shootings = length(STATISTICAL_MURDER_FLAG)) %>%
  select(BORO, year, murders, shootings) %>%
  ungroup()
```

Second visualization: Murders and shootings by year for Bronx 
```{r visualize_New_York}
boro <- "BRONX"
boro_group %>% 
  filter(BORO == boro) %>% 
  ggplot(aes(x = year, y = murders)) + 
  geom_line(aes(color = "murders")) + 
  geom_point(aes(color = "murders")) + 
  geom_line(aes(y = shootings, color = "shootings")) + 
  geom_point(aes(y = shootings, color = "shootings")) + 
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = str_c("Murders & Shootings in ", boro), y = NULL)
```

Find the three boroughs with the highest murder counts since 2010
```{r worst_boro}
boro_group %>% filter(year >= 2010) %>% group_by(BORO) %>% summarize(murders = sum(murders)) %>% slice_max(murders, n = 3)
```

Bias identification: At first I was very interested in seeing how race and age might play out in these shooting incidents, 
but then realized how fraught with biases both of these were, both my own and in the race identifications available in the data as well as the very broad age groupings that were used.

So to avoid these biases both my own and in the data, I looked only at murders and shootings as they relate to time, either month of the year or year over year.  The exception to this is the analysis of the boroughs with the highest number of murders.  One might think that Manhattan is a safer place from this, but instead it could be that most murders happen in the evenings and Manhattan has more businesses than residences. To find out if this could be biasing the results would require further research and data.